---
title: "Assignment4_Clustering"
author: "Bodduluri Yashasree"
date: "2023-11-10"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(factoextra) # clustering algorithms & visualization
library(ISLR)
library(caret)
```

#Importing the dataset
```{r}
Pharmaceuticals <- read.csv("C:\\Users\\HP\\Desktop\\Pharmaceuticals.csv")
summary(Pharmaceuticals)
```

# a. Cluster the 21 firms using only the numerical variables (1–9). Justify the various decisions made during the cluster analysis, such as variable weights, the specific clustering algorithm(s) used, the number of clusters formed, and so on.

#Before clustering data, remove missing data and rescale variables for comparability.

```{r}
Pharma<- na.omit(Pharmaceuticals) #gives the data after removing the missing values.
Pharma
```

#Only the quantitative variables (1-9) are required to cluster the 21 firms.
```{r}
row.names(Pharma)<- Pharma[,1]
Pharma_1<- Pharma[,3:11]
head(Pharma_1)
```

#Scale all the dataframe's quantitative variables
```{r}
Pharma_2<-scale(Pharma_1)
head(Pharma_2)
```

##Using the Elbow Method, determine the number of clusters to be used in the cluster analysis.
```{r}
fviz_nbclust(Pharma_2, kmeans, method = "wss") + labs(subtitle = "Elbow Method")
```

##Using the Silhouette method to determine the number of clusters

```{r}
fviz_nbclust(Pharma_2, kmeans, method = "silhouette")+ labs(subtitle = "Silhouette Method")
```

The number of clusters is 5 in the above plots, which is sufficient to display the data 
variations.

```{r}
set.seed(64060)
k5<- kmeans(Pharma_2,centers=5,nstart = 25) 
```

#Visualizing the output
```{r}
k5$centers  #for centroids
fviz_cluster(k5,data = Pharma_2) # to Visualize the clusters
k5
distance<- dist(Pharma_2, method = "euclidean")
fviz_dist(distance)
```
#K-Means Cluster Analysis was used to fit the data into 5 clusters.

```{r}
fit<-kmeans(Pharma_2,5)
```

#calculating the mean of all quantitative variables in each cluster
```{r}
aggregate(Pharma_2,by=list(fit$cluster),FUN=mean)
```

```{r}
Pharma_3<-data.frame(Pharma_2,fit$cluster)
Pharma_3
```

#view of the cluster plot
```{r}
library(cluster)
clusplot(Pharma_2,fit$cluster,color = TRUE,shade = TRUE,labels = 2,lines = 0)
```


#b.Interpret the clusters in relation to the numerical variables that were used to form the clusters.
By examining the mean values of all quantitative variables within each cluster.

Cluster 1 consists of JNJ, MRK, PFE, and GSK. Cluster 1 has the highest Market_cap,ROA,ROE,Asset_Turnover, and the lowest Beta,PE_Ratio.

Cluster 2 has the highest Rev_Growth and the lowest PE_Ratio, Asset_Turnover.

Cluster 3 has the highest Beta, Leverage, and the lowest Market_Cap, ROE, ROA, Leverage, Rev_Growth, and Net_Profit_Margin.

Cluster 4 has the highest PE_Ratio and the lowest Leverage and Asset_Turnover.

Cluster 5 consists of the following stocks: AZN,ABT,NVS,BMY,WYE,SGP,LLY. Cluster 5 has the highest Net_Profit_Margin and the lowest leverage, Beta.
 
# c.s there a pattern in the clusters with respect to the numerical variables (10 to 12)? (those not used in forming the clusters)  
There is a pattern in the clusters for the Media recommendation variable.

Cluster 1 has an equal Hold and Moderate Buy recommendation due to its highest Market_Cap, highest ROE, highest ROA, and highest Asset_Turnover.

Hold is recommended for Cluster 2, which has the lowest PE_Ratio and Asset_Turnover.

Cluster-3, which has the highest Beta and Leverage, has mostly Moderate Buy Recommendation.

Hold is recommended for Cluster 4, which has the highest PE_Ratio.

Cluster 5 has the highest Net_Profit_Margin and is mostly recommended to hold.

In terms of variables, I've noticed a trend among the clusters (10 to 12).

Clusters 1–3 have a majority of Moderate Buy Recommendation.

Clusters 1,2,4,5 are recommended to be held.


# d.Provide an appropriate name for each cluster using any or all of the variables in the dataset. 
Cluster-1 - Buy (or) Hold Moderately.

Cluster-2 - Low PE_Ratio, Asset_Turnover, or Hold.

Cluster-3 - High Beta, Buy Cluster (or Leverage Cluster).

Cluster-4 is a high PE_Ratio (or high Hold) cluster.

Cluster-5 is a high net profit margin (or high hold) cluster.

